|

### 2020-06-02 paper/Manuscript_JSS.Rmd
Prepare manuscript for JSS submission

---
title: |
    | WORCS: A Workflow for Open Reproducible Code in Science
    | `r paste0("Version: ", worcs:::cran_version(), ", date: ", format(Sys.Date(), "%d %b,\ %Y"))`

shorttitle        : "WORKFLOW FOR OPEN SCIENCE"

author:
  - name: "Caspar J. Van Lissa"
    affiliation: "1,2"
    corresponding: yes
    address: "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email: "c.j.vanlissa@uu.nl"
  - name: "Andreas M. Brandmaier"
    affiliation: "3,4"
  - name: "Loek Brinkman"
    affiliation: "2,5"
  - name: "Anna-Lena Lamprecht"
    affiliation: "2,6"
    email: "a.l.lamprecht@uu.nl"
  - name: "Aaron Peikert"
    affiliation: "3,7"
  - name: "Marijn E. Struiksma"
    affiliation: "2,8"
  - name: "Barbara Vreede"
    affiliation: "9"

affiliation:
  - id: "1"
    institution: "Utrecht University faculty of Social and Behavioral Sciences, department of Methodology & Statistics"
  - id: "2"
    institution: "Open Science Community Utrecht"
  - id: "3"
    institution: "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id: "4"
    institution: "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany and London, UK"
  - id: "5"
    institution: "University Medical Center Utrecht"
  - id: "6"
    institution: "Utrecht University, Department of Information and Computing Sciences"
  - id: "7"
    institution: "Humboldt-Universitaet zu Berlin"
  - id: "8"
    institution: "Utrecht Institute of Linguistics OTS"
  - id: "9"
    institution: "University Library Utrecht"

abstract: |
  Adopting open science
  principles can be challenging and time-intensive, because doing so requires
  substantial conceptual education and training in the use of new tools. The 
  goal of this paper is to lower the threshold to adoption. We introduce the Workflow for
  Open Reproducible Code in Science (WORCS). WORCS constitutes a step-by-step
  procedure that researchers can follow to make a research project open and
  reproducible. It is based on best practices, and can be used either in
  parallel to, or in absence of, top-down requirements by journals,
  institutions, and funding bodies. This workflow is primarily targeted towards
  scholars working in R on a private computer,
  conducting research that involves academic prose, analysis code,
  and (optionally) tabular data - but it is flexible enough to accommodate many
  other scenarios.
  To facilitate widespread adoption, the
  WORCS principles are implemented in the R package `worcs`, which offers an
  RStudio 
  project template that is, effectively, a "one-click solution" for R users. 
  This paper introduces the workflow, discusses how it meets different 
  requirements of open science, and provides examples of the implementation of
  `worcs` in R.

authornote: |
  This is a preprint, not a peer-reviewed publication. Comments and
  contributions are welcome. Please contact the lead author.
  All authors of this paper, except the first, are listed in alphabetical order.
  Author contributions are detailed in the document 'authors.csv' on the project
  repository, at https://github.com/cjvanlissa/worcs.
  The lead author is supported by a NWO Veni grant (NWO grant number
  VI.Veni.191G.090).
  We acknowledge Jeroen Ooms for offering feedback and suggesting the use of the
  R package gert.
keywords          : "open science; reproducibility; r; dynamic document generation; version control"
wordcount         : "8257"

bibliography      : ["worcs.bib"]

figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes
mask              : no

class             : "man"
classoption       : "noextraspace"
output            : papaja::apa6_pdf
urlcolor          : blue
knit              : worcs::cite_all
---

```{r load_packages, include = FALSE}
library("papaja")
mask <- TRUE
masked <- function(text){
  ifelse(mask, "[masked]", text)
}
```

```{r analysis_preferences}
# Seed for random number generation
set.seed(42)
```

\setlength{\parskip}{0em}

<!--\vspace{\baselineskip}-->

@@

Many researchers are motivated to adopt current best practices for open
science and enjoy these benefits. And yet, the question of "how" to 
apply open science practices can be daunting. Making the transition
requires researchers to become knowledgeable about different open science 
challenges and solutions, and to become proficient with new and unfamiliar tools.
This paper is designed to ease that transition by presenting a simple workflow 
that meets most requirements open science, and is based on best practices:
The *Workflow for Open Reproducible Code in Science* (WORCS).
WORCS is a lightweight approach to open science and
computational reproducibility for scholars working in R on a private computer,
conducting research that involves academic prose, analysis code,
and (optionally) tabular data - but it is flexible enough to accommodate many
other scenarios. The workflow (Figure \@ref(fig:workflow)) 
constitutes a step-by-step procedure that researchers can follow to make a 
research project (including data, code, and a report) open and reproducible.
WORCS is compatible with open science
requirements already implemented by journals and institutions, and will help 
fulfill them. It can also be used in the absence of top-down support or guidelines
to produce work in accordance with best practices.
Although the principles underlying this workflow are universal, WORCS has first
been implemented for R users [@rcoreteamLanguageEnvironmentStatistical2020].
The R package `worcs` [@worcspackage2020] offers a project template for RStudio
[@rstudioteamRStudioIntegratedDevelopment2015], which is, effectively,
a "one click solution". This paper introduces the WORCS workflow, discusses how
it meets best practices for open science, and illustrates the use of the `worcs`
package.

```{r workflow, fig.cap="Schematic illustration of the WORCS procedure", out.width="100%"}
knitr::include_graphics("workflow_graph/workflow.png")
```

@@

In the R-implementation of WORCS, we recommend centering a research project
around one dynamically generated 'RMarkdown' document, which includes all 
analysis code. Longer scripts can be
stored in `.R` files, and called from the main document using the `source()`
function. When a reader or reviewer compiles this document, all code is run
automatically, thus verifying computational replicability. The 'RMarkdown' 
document can be automatically formatted in many styles, including
APA style [thanks to the R package `papaja`, @austPapajaPrepareReproducible2020],
a host of other scientific formats 
[see @allaireRticlesArticleFormats2020], and as plain 'Markdown'.

@@

An integral part of WORCS is the *explicit* version control software 'Git'
(www.git-scm.com). Git tracks
changes to files, and stores these changes when the user makes a "commit" (a 
snapshot of the version controlled files).
Git retains a complete log of all commits, and users can compare changes between
different commits, or go back to a previous version of the code (for example,
after making a mistake, or to replicate a previous version of the results).
A project version controlled with Git is called a "Repository", or Repo.
Git only version controls files explicitly committed by the user. Moreover, it
is possible to prevent files from being version controlled - which is useful for
privacy sensitive data.
A `.gitignore` file lists files that should not be version-controlled.
These files thus exist only on the user's private computer.

The functionality of 'Git' is amplified by services such as 'GitHub'
(https://github.com). 'GitHub' is best understood as a cloud storage service
with social networking functionality. The cloud storage aspect
of 'GitHub' works as follows: You can "clone" (copy) a local 'Git' repository to
the 'GitHub' website, as a backup or research archive. The social network aspect
comes into play when a repository on 'GitHub' is set to "public": This allows 
other researchers to peruse the repository and see how the work was done; clone
it to their own computer to replicate the original work or apply the methods to
their own data; open "Issues" to ask questions or give feedback on the project,
or even send a "Pull request" with suggested changes to the text or code for
your consideration. 'Git' and 'GitHub' shine as tools for collaboration, because
different people can simultaneously work on different parts of a project, and
their changes can be compared and automatically merged on the website. Even on
solo projects, working with 'Git'/'GitHub' has many benefits: Staying organized,
being able to start a new study with a clone of an old, similar repository, or
splitting off an "experimental branch" to try something new, while retaining the
ability to "revert" (return) to a previous state of the project, or to "merge"
(incorporate) the experimental branch.

It is worth noting that,
although this paper focuses on the most widely used 'Git'
remote repository service, 'GitHub', two notable alternatives exist: 
'GitLab' (https://gitlab.com) and 'Bitbucket' (https://bitbucket.org).
In 
order to present the workflow in a platform-agnostic way, we will refer to such
cloud hosting services as "remote repositories" in the remainder of this
paper.
As of
version 0.1.2, `worcs` has been tested with these three remote repository services,
but has been optimized for use with 'GitHub'.
A [vignette](https://cjvanlissa.github.io/worcs/articles/git_cloud.html) is included
with instructions for using alternative platforms. 
When initializing a new project using the `worcs` project template,
a 'https' remote repository address will usually work best with the authentication procedure outlined in the
[setup vignette](https://cjvanlissa.github.io/worcs/articles/setup.html).

The R-implementation of WORCS uses the package `gert`, a 'Git' client for 'R'
[@oomsGertSimpleGit2019], to connect a local project to a remote repository.
The `worcs` package also contains a user-friendly function to set
the default user credentials, `git_user()`, and a function to add, commit, and
push all changed files to a remote repository in one step, `git_update()`.

@@

The third solution is *dependency management*: Keeping track of exactly what
software was used to conduct the analyses. At first glance, it might seem
sufficient to state that analyses were conducted in *Program X*. However, every
program is susceptible to changes, updates, and bugfixes. A lot of Open Source software,
in particular, is regularly updated because there is an
active community of developers contributing functionality and bugfixes. 
Potentially, any such update could change the results of the code, thus rendering
the analysis 'computationally non-reproducible'.

Many solutions exist to ensure computational reproducibility. These solutions typically work by
enveloping your research project in a distinct "environment" that only has access to programs that
are explicitly installed, and maintaining a record of these programs. These solutions differ in
user-friendliness and effectiveness. If strict reproducibility is required, users might choose to conduct all analyses in a 'Docker container': An environment that behaves like a
virtual computer, which can be stored like a sort of time capsule, and
identically reinstated on a user computer, or in the cloud.
This is effective, and also preserves software outside the R environment (e.g.,
preprocessing tools for neuroimaging data). A Docker-based workflow
was recently introduced by @peikertReproducibleDataAnalysis2019. This workflow
is based on many of the same principles as WORCS, and the two approaches are
conceptually and practically compatible. However, the workflow described by 
Peikert and Brandmaier can be difficult to set up for novice users. 
The accompanying R package `repro`
is currently in development to facilitate this transition [@repro]. Another 
user-friendly Docker-based workflow is offered by the cloud-based collaboration
platform 'Digital Ocean'. Although Digital Ocean enjoys some popularity among 
scientists, it is not clear whether this platform's
["Data Processing Agreement"](https://www.digitalocean.com/legal/data-processing-agreement/)
is GDPR compliant - it seems to focus more on limiting liability than on
compliance. The necessity to upload data for analysis is a risk with any
cloud-computing solution when human participants are involved.

For analyses conducted entirely in R, the recently released package `renv`
strikes a great balance between user-friendliness  and effectiveness. Developed by
the team behind 'RStudio', this package simply maintains a text-based,
human-readable log of all packages used, their version numbers, and where they
were installed from (e.g., 'CRAN', 'Bioconductor', 'GitHub'). As long as these
repositories maintain their archives, this solution is reliable and very
lightweight.
The text-based log file can be version controlled with 'Git'. `renv`
installs the logged packages in a cache that is shared between all
projects on the computer. Thus, if one of your projects uses a package that is
already in the cache, it does not
need to be installed again. This prevents long installation
times, large space requirements, and frustration having to re-install software
that is already installed on the system. When someone else loads your 
project, `renv` will install all of the required packages onto a cache on their
computer. 
This is not a foolproof method to ensure strict computational reproducibility, but it is more user-friendly
and lightweight than any alternative out there, requiring no technical skill to
setup. When choosing an appropriate solution for dependency management, it is
important to consider that all solutions have a limited shelf life
[@brownHowLearnedStop2017].
Ultimately, the best approach might be to use a "just good-enough" solution, and
acknowledge that all code requires some maintenance if you want to
reproduce it in the future. For projects in R, `renv` meets these requirements.

@@

A key consideration when developing a research project is what filetypes to use.
A case can be made to use text-based files exclusively (or as much as possible),
instead of binary files.
Text-based files can be read by machines and humans alike.
Binary files, such as 'Word' (`.docx`) or 'SPSS' (`.sav`, `.spo`) files,
must be decoded first, and many require commercial software to do so.
The version control software 'Git' is also designed for use with text-based files.
The 'Git' change log reflects changes to text-based files on a line-by-line basis.
Binary files can be version controlled with 'Git', but 
the change log for a binary file are uninterpretable.
Binary files are also often 
larger than text-based files, which means they take up more space on cloud
hosting services. For these reasons, uploading large binary
files to remote repositories is frowned upon. 
WORCS encourages the use of
text-based files for all purposes.

Two additional points are worth noting:
First, 'Git' tracks line-by-line changes to text-based files,
based on line breaks in the text.
When writing a version-controlled paper in plain-text,
it is therefore recommended to start each sentence on a new line.
That way, the change log will indicate which specific sentence was edited, 
instead of replacing an entire paragraph.
As RMarkdown does not parse a single line break as the end of a paragraph,
this practice does not disrupt the flow of a paragraph in the rendered version.
Second, the remote repository service 'GitHub' rewards the use of text-based files by
rendering certain text-based filetypes for online viewing: 
For example, `.md`
(Markdown) files are displayed as web pages, and `.csv` files as spreadsheets.
The `worcs` implementation for R makes use of this fact by, for example,
rendering a codebook for the data as `.md` so visitors of a 'Git' remote repository
can view it as a web page.

@@

We provide a conceptual outline of the WORCS procedure below, based on WORCS
Version 0.1.2. 
For users of the `worcs` R package, we highly recommend referring to
[the workflow vignette](https://cjvanlissa.github.io/worcs/articles/setup.html),
which is more detailed and updated with future developments.

@@

## The R-implementation of WORCS

The R package `worcs` automates most steps of the workflow
for R users.
Several arguments support the choice to implement this workflow in R.
First, R and all of it extensions are free open source software, which
make it a tool of choice for open science.
Second, all tools required for an open science workflow are implemented in R,
most of these tools are directly accessible through the user interface of RStudio,
and some of them are actively developed by the team behind RStudio.
Third, R is the second-most cited statistical software package [@muenchenPopularityDataScience2012], following SPSS, which has no support for any of the open science tools discussed in this paper.
Fourth, R is well-supported;
a vibrant and inclusive online community exists for R,
developing new R methods and packages,
and providing support and tutorials for existing ones.
Finally, R is highly interoperable:
Packages are available to load nearly every imaginable filetype, and output can be written to most file types, including DOCX and PDF (in APA style), and HTML format. 
Moreover, wrappers are available for many tools developed in other programming languages, and code written in other programming languages can be evaluated from R, including C++, 
Fortran, and Python [@allaireMarkdownPythonEngine2020]. There are excellent free
resources for learning R, such as the book 'R for Data Science' [@grolemundDataScience2017].

Working with R is simplified immensely by using the RStudio integrated
development engine (IDE) for R [@rstudioteamRStudioIntegratedDevelopment2015].
RStudio automates and streamlines tedious or complicated aspects of working with
the tools used in WORCS, and many are embedded directly into the visual user
interface.
This makes RStudio a comprehensive solution for open science research projects.
Another important feature of RStudio is project management.
A project bundles writing, analyses, data, references, et cetera, into a
self-contained folder, that can be uploaded entirely to a 'Git' remote repository and downloaded
by future users. The `worcs` R package installs a new RStudio project template.
When a new project is initialized from this template, the bookkeeping required
to set up an open science project is performed automatically.

@@

Before you can use the R-implementation of the WORCS workflow, you have to install the required software.
You only have to perform this setup procedure only once for each computer you
intend to use `worcs` on. It should take about 30 minutes, and it is documented
in [the setup vignette](https://cjvanlissa.github.io/worcs/articles/setup.html).
The setup vignette does not assume a prior installation of R, so it is suitable for novice users.
After setting up your system, you can use `worcs` for all of your
projects. The R-implementation of the workflow introduced earlier is detailed,
step-by-step, in
[the workflow vignette](https://cjvanlissa.github.io/worcs/articles/workflow.html).

@@

References for software are sometimes provided within the software environment;
for example, in R, package references can be found by
calling `citation("packagename")`. This returns an APA-style reference, and a
BibTeX entry. Software papers are also common, and are sometimes called
"Application Notes". The [Journal of Open Source Software (JOSS)](https://joss.theoj.org/)
offers programmers a way to generate a publication 
based on well-documented software;
in this process, the code itself is peer reviewed, and writing an actual paper is optional. For an example, see
@rosenbergTidyLPAPackageEasily2018.
Software citations can also sometimes be found 
in the source code repository (e.g., on 'Git' remote repositories), using the
['Citation File Format'](https://citation-file-format.github.io/). The
[*Software Heritage Project*](https://www.softwareheritage.org/2020/05/26/citing-software-with-style/)
recently published a BibLaTeX software citation style that will further facilitate
accurate citation of software in dynamically generated documents.

@@

## Implementation in `worcs`

The `worcs` package includes a [vignette on citation](https://cjvanlissa.github.io/worcs/articles/citing.html) to explain the
fundamentals for novice users, and offer recommendations specific to `worcs`. We
recommend using the free, open-source reference manager
[Zotero](https://www.zotero.org/download/); it is feature-rich, 
user-friendly, and highly interoperable with other reference managers. A 
tutorial for using Zotero with RMarkdown exists  [here](https://christopherjunk.netlify.com/blog/2019/02/25/zotero-RMarkdown/).

The `worcs` package also offers original solutions for comprehensive citation.
Firstly,
during the writing process, authors can mark the distinction between essential
and non-essential references. It is easier to do this from the start, instead of
going back to cut non-essential references just prior to publication. Standard
Markdown uses the at-symbol (`@`) to cite a reference. `worcs` additionally 
reserves the "double at"-symbol (`@@`) to cite a non-essential reference.
Users can render the manuscript either with, or without, comprehensive citations
by adapting the `knit` command in the front matter, setting it to
`knit: worcs::cite_all` to render all citations, and to 
`knit: worcs::cite_essential` to remove all *non-essential* citations.

With regard to the citation of R packages, it is worth noting that the `papaja` 
package [@austPapajaPrepareReproducible2020], which provides the 'APA6'
template used in `worcs`, also
includes functions to automatically cite all packages used in the R-session, and
add their references to the bibliography file.

@@

WORCS offers a workflow for open reproducible code in science. The step-by-step procedure outlined in this tutorial helps researchers make an entire research project Open and Reproducible. The accompanying R package provides user-friendly support functions for several steps in the workflow, and an RStudio project template to get the project set up correctly.

>>>>

---
documentclass: jss
author:
  - name: Caspar J. van Lissa
    affiliation: 'Utrecht University'
    address: Padualaan 14, 3584CH Utrecht, The Netherlands
    email: c.j.vanlissa@uu.nl
  - name: Andreas M. Brandmaier
    affiliation: 'Max Planck Institute'
  - name: Loek Brinkman
    affiliation: 'UMC Utrecht \AND'
  - name: Anna-Lena Lamprecht
    affiliation: 'Utrecht University'
    email: a.l.lamprecht@uu.nl
  - name: Aaron Peikert
    affiliation: Max Planck Institute
  - name: Marijn E. Struiksma
    affiliation: 'Utrecht University'
  - name: Barbara Vreede
    affiliation: 'Utrecht University'
title:
  formatted: >
    WORCS: A Workflow for Open Reproducible Code in Science
  short:      "WORKFLOW FOR OPEN SCIENCE"
abstract: >
  Adopting open science
  principles can be challenging and time-intensive, because doing so requires
  substantial conceptual education and training in the use of new tools.
  This paper introduces the Workflow for
  Open Reproducible Code in Science (WORCS): A step-by-step
  procedure that researchers can follow to make a research project open and
  reproducible, accompanied by an open source software implementation that 
  automates these steps.
  The 
  purpose of the workflow is to lower the threshold for 
  adoption of open science principles. 
  It is based on best practices, and can be used either in
  parallel to, or in absence of, top-down requirements by journals,
  institutions, and funding bodies.
  This workflow is primarily targeted towards
  scholars working in R on a private computer,
  conducting research that involves academic prose, analysis code,
  and (optionally) tabular data - but it is flexible enough to accommodate many
  other scenarios.
  To facilitate widespread adoption, the
  WORCS principles are implemented in the R package \pkg{worcs}, which offers an
  RStudio 
  project template that is, effectively, a "one-click solution" for R users,
  and utility functions for specific workflow steps.
  This paper addresses the design of the workflow,
  discusses how it meets different 
  standards for open science, and provides examples of the implementation of
  \pkg{worcs} in \proglang{R}.
authornote: "blabla"
keywords:
  # at least one keyword must be supplied
  formatted: [open science, reproducibility, r, dynamic document generation, version control]
  plain:     [open science, reproducibility, r, dynamic document generation, version control]
preamble: >
  \usepackage{amsmath}
bibliography      : ["worcs.bib"]  
urlcolor          : blue
output: rticles::jss_article
knit: worcs::cite_all
---

```{r load_packages, include = FALSE}
library("papaja")
mask <- TRUE
masked <- function(text){
  ifelse(mask, "[masked]", text)
}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE, warning = FALSE, message= FALSE)
```

@@

Many researchers are motivated to adopt current best practices for open
science and enjoy these benefits. And yet, the question of "how" to 
apply open science practices can be daunting. Making the transition
requires researchers to become knowledgeable about different open science 
challenges and solutions, and to become proficient with new and unfamiliar tools.
This paper is designed to ease that transition by presenting a simple workflow 
that meets most requirements open science, and is based on best practices:
The *Workflow for Open Reproducible Code in Science* (WORCS).
WORCS is a lightweight approach to open science and
computational reproducibility for scholars working in \proglang{R} on a private computer,
conducting research that involves academic prose, analysis code,
and (optionally) tabular data - but it is flexible enough to accommodate many
other scenarios. The workflow (Figure \@ref(fig:workflow)) 
constitutes a step-by-step procedure that researchers can follow to make a 
research project (including data, code, and a report) open and reproducible.
WORCS is compatible with open science
requirements already implemented by journals and institutions, and will help 
fulfill them. It can also be used in the absence of top-down support or guidelines
to produce work in accordance with best practices.
Although the principles underlying this workflow are universal, WORCS has first
been implemented for \proglang{R} users [@rcoreteamLanguageEnvironmentStatistical2020].
The \proglang{R} package \pkg{worcs} [@worcspackage2020] offers a project template for RStudio
[@rstudioteamRStudioIntegratedDevelopment2015], which is, effectively,
a "one click solution". This paper introduces the WORCS workflow, discusses how
it meets best practices for open science, and illustrates the use of the \pkg{worcs}
package.

```{r workflow, fig.cap="Schematic illustration of the WORCS procedure.", out.width="100%"}
#knitr::include_graphics("workflow_graph/workflow.png")
knitr::include_graphics("workflow_graph/workflow.svg")
```

@@

In the \proglang{R} implementation of WORCS, we recommend centering a research project
around one dynamically generated \code{RMarkdown} document, which includes all 
analysis code. Longer scripts can be
stored in \code{.R} files, and called from the main document using the \code{source()}
function. When a reader or reviewer compiles this document, all code is run
automatically, thus verifying computational replicability. The \code{RMarkdown} 
document can be automatically formatted in many styles, including
APA style [thanks to the \proglang{R} package \pkg{papaja}, @austPapajaPrepareReproducible2020],
a host of other scientific formats 
[see @allaireRticlesArticleFormats2020], and as plain \code{Markdown}.

@@

An integral part of WORCS is the *explicit* version control software Git
(www.git-scm.com). Git tracks
changes to files, and stores these changes when the user makes a "commit" (a 
snapshot of the version controlled files).
Git retains a complete log of all commits, and users can compare changes between
different commits, or go back to a previous version of the code (for example,
after making a mistake, or to replicate a previous version of the results).
A project version controlled with Git is called a "Repository", or Repo.
Git only version controls files explicitly committed by the user. Moreover, it
is possible to prevent files from being version controlled - which is useful for
privacy sensitive data.
A \code{.gitignore} file lists files that should not be version-controlled.
These files thus exist only on the user's private computer.

The functionality of Git is amplified by services such as GitHub
(https://github.com). GitHub is best understood as a cloud storage service
with social networking functionality. The cloud storage aspect
of GitHub works as follows: You can "clone" (copy) a local Git repository to
the GitHub website, as a backup or research archive. The social network aspect
comes into play when a repository on GitHub is set to "public": This allows 
other researchers to peruse the repository and see how the work was done; clone
it to their own computer to replicate the original work or apply the methods to
their own data; open "Issues" to ask questions or give feedback on the project,
or even send a "Pull request" with suggested changes to the text or code for
your consideration. Git and GitHub shine as tools for collaboration, because
different people can simultaneously work on different parts of a project, and
their changes can be compared and automatically merged on the website. Even on
solo projects, working with Git/GitHub has many benefits: Staying organized,
being able to start a new study with a clone of an old, similar repository, or
splitting off an "experimental branch" to try something new, while retaining the
ability to "revert" (return) to a previous state of the project, or to "merge"
(incorporate) the experimental branch.

It is worth noting that,
although this paper focuses on the most widely used Git
remote repository service, GitHub, two notable alternatives exist: 
GitLab (https://gitlab.com) and Bitbucket (https://bitbucket.org).
In 
order to present the workflow in a platform-agnostic way, we will refer to such
cloud hosting services as "remote repositories" in the remainder of this
paper.
As of
version 0.1.2, \pkg{worcs} has been tested with these three remote repository services,
but has been optimized for use with GitHub.
A [vignette](https://cjvanlissa.github.io/worcs/articles/git_cloud.html) is included
with instructions for using alternative platforms. 
When initializing a new project using the \pkg{worcs} project template,
a 'https' remote repository address will usually work best with the authentication procedure outlined in the
[setup vignette](https://cjvanlissa.github.io/worcs/articles/setup.html).

The \proglang{R} implementation of WORCS uses the package \code{gert}, a Git client for \proglang{R}
[@oomsGertSimpleGit2019], to connect a local project to a remote repository.
The \pkg{worcs} package also contains a user-friendly function to set
the default user credentials, \code{git_user()}, and a function to add, commit, and
push all changed files to a remote repository in one step, \code{git_update()}.

@@

The third solution is *dependency management*: Keeping track of exactly what
software was used to conduct the analyses. At first glance, it might seem
sufficient to state that analyses were conducted in *Program X*. However, every
program is susceptible to changes, updates, and bugfixes. A lot of Open Source software,
in particular, is regularly updated because there is an
active community of developers contributing functionality and bugfixes. 
Potentially, any such update could change the results of the code, thus rendering
the analysis computationally non-reproducible.

Many solutions exist to ensure computational reproducibility. These solutions typically work by
enveloping your research project in a distinct "environment" that only has access to programs that
are explicitly installed, and maintaining a record of these programs. These solutions differ in
user-friendliness and effectiveness. If strict reproducibility is required, users might choose to conduct all analyses in a "Docker container": An environment that behaves like a
virtual computer, which can be stored like a sort of time capsule, and
identically reinstated on a user computer, or in the cloud.
This is effective, and also preserves software outside the \proglang{R} environment (e.g.,
preprocessing tools for neuroimaging data). A Docker-based workflow
was recently introduced by @peikertReproducibleDataAnalysis2019. This workflow
is based on many of the same principles as WORCS, and the two approaches are
conceptually and practically compatible. However, the workflow described by 
Peikert and Brandmaier can be difficult to set up for novice users. 
The accompanying \proglang{R} package \pkg{repro}
is currently in development to facilitate this transition [@repro]. Another 
user-friendly Docker-based workflow is offered by the cloud-based collaboration
platform Digital Ocean. Although Digital Ocean enjoys some popularity among 
scientists, it is not clear whether this platform's
["Data Processing Agreement"](https://www.digitalocean.com/legal/data-processing-agreement/)
is GDPR compliant - it seems to focus more on limiting liability than on
compliance. The necessity to upload data for analysis is a risk with any
cloud-computing solution when human participants are involved.

For analyses conducted entirely in \proglang{R}, the recently released package \pkg{renv}
strikes a great balance between user-friendliness  and effectiveness. Developed by
the team behind RStudio, this package simply maintains a text-based,
human-readable log of all packages used, their version numbers, and where they
were installed from (e.g., CRAN, Bioconductor, GitHub). As long as these
repositories maintain their archives, this solution is reliable and very
lightweight.
The text-based log file can be version controlled with Git. \pkg{renv}
installs the logged packages in a cache that is shared between all
projects on the computer. Thus, if one of your projects uses a package that is
already in the cache, it does not
need to be installed again. This prevents long installation
times, large space requirements, and frustration having to re-install software
that is already installed on the system. When someone else loads your 
project, \pkg{renv} will install all of the required packages onto a cache on their
computer. 
This is not a foolproof method to ensure strict computational reproducibility, but it is more user-friendly
and lightweight than any alternative out there, requiring no technical skill to
setup. When choosing an appropriate solution for dependency management, it is
important to consider that all solutions have a limited shelf life
[@brownHowLearnedStop2017].
Ultimately, the best approach might be to use a "just good-enough" solution, and
acknowledge that all code requires some maintenance if you want to
reproduce it in the future. For projects in \proglang{R}, \pkg{renv} meets these requirements.

@@

A key consideration when developing a research project is what filetypes to use.
A case can be made to use text-based files exclusively (or as much as possible),
instead of binary files.
Text-based files can be read by machines and humans alike.
Binary files, such as Word (\code{.docx}) or SPSS (\code{.sav}, \code{.spo}) files,
must be decoded first, and many require commercial software to do so.
The version control software Git is also designed for use with text-based files.
The Git change log reflects changes to text-based files on a line-by-line basis.
Binary files can be version controlled with Git, but 
the change log for a binary file are uninterpretable.
Binary files are also often 
larger than text-based files, which means they take up more space on cloud
hosting services. For these reasons, uploading large binary
files to remote repositories is frowned upon. 
WORCS encourages the use of
text-based files for all purposes.

Two additional points are worth noting:
First, Git tracks line-by-line changes to text-based files,
based on line breaks in the text.
When writing a version-controlled paper in plain-text,
it is therefore recommended to start each sentence on a new line.
That way, the change log will indicate which specific sentence was edited, 
instead of replacing an entire paragraph.
As RMarkdown does not parse a single line break as the end of a paragraph,
this practice does not disrupt the flow of a paragraph in the rendered version.
Second, the remote repository service GitHub rewards the use of text-based files by
rendering certain text-based filetypes for online viewing: 
For example, \code{.md}
(Markdown) files are displayed as web pages, and \code{.csv} files as spreadsheets.
The \pkg{worcs} implementation for \proglang{R} makes use of this fact by, for example,
rendering a codebook for the data as \code{.md} so visitors of a Git remote repository
can view it as a web page.

@@

We provide a conceptual outline of the WORCS procedure below, based on WORCS
Version 0.1.2. 
For users of the \pkg{worcs} \proglang{R} package, we highly recommend referring to
[the workflow vignette](https://cjvanlissa.github.io/worcs/articles/setup.html),
which is more detailed and updated with future developments.

@@

## The R implementation of WORCS

The \proglang{R} package \pkg{worcs} automates most steps of the workflow
for \proglang{R} users.
Several arguments support the choice to implement this workflow in \proglang{R}.
First, \proglang{R} and all of it extensions are free open source software, which
make it a tool of choice for open science.
Second, all tools required for an open science workflow are implemented in \proglang{R},
most of these tools are directly accessible through the user interface of RStudio,
and some of them are actively developed by the team behind RStudio.
Third, \proglang{R} is the second-most cited statistical software package [@muenchenPopularityDataScience2012], following SPSS, which has no support for any of the open science tools discussed in this paper.
Fourth, \proglang{R} is well-supported;
a vibrant and inclusive online community exists for \proglang{R},
developing new \proglang{R} methods and packages,
and providing support and tutorials for existing ones.
Finally, \proglang{R} is highly interoperable:
Packages are available to load nearly every imaginable filetype, and output can be written to most file types, including DOCX and PDF (in APA style), and HTML format. 
Moreover, wrappers are available for many tools developed in other programming languages, and code written in other programming languages can be evaluated from \proglang{R}, including \proglang{C++}, 
\proglang{Fortran}, and \proglang{Python} [@allaireMarkdownPythonEngine2020]. There are excellent free
resources for learning \proglang{R} [e.g., @grolemundDataScience2017].

Working with \proglang{R} is simplified immensely by using the RStudio integrated
development engine (IDE) for \proglang{R} [@rstudioteamRStudioIntegratedDevelopment2015].
RStudio automates and streamlines tedious or complicated aspects of working with
the tools used in WORCS, and many are embedded directly into the visual user
interface.
This makes RStudio a comprehensive solution for open science research projects.
Another important feature of RStudio is project management.
A project bundles writing, analyses, data, references, et cetera, into a
self-contained folder, that can be uploaded entirely to a Git remote repository and downloaded
by future users. The \pkg{worcs} \proglang{R} package installs a new RStudio project template.
When a new project is initialized from this template, the bookkeeping required
to set up an open science project is performed automatically.

@@

Before you can use the \proglang{R} implementation of the WORCS workflow, you have to install the required software.
You only have to perform this setup procedure only once for each computer you
intend to use \pkg{worcs} on. It should take about 30 minutes, and it is documented
in [the setup vignette](https://cjvanlissa.github.io/worcs/articles/setup.html).
The setup vignette does not assume a prior installation of \proglang{R}, so it is suitable for novice users.
After setting up your system, you can use \pkg{worcs} for all of your
projects. The \proglang{R} implementation of the workflow introduced earlier is detailed,
step-by-step, in
[the workflow vignette](https://cjvanlissa.github.io/worcs/articles/workflow.html).

@@

References for software are sometimes provided within the software environment;
for example, in \proglang{R}, package references can be found by
calling \code{citation("packagename")}. This returns an APA-style reference, and a
BibTeX entry. Software papers are also common, and are sometimes called
"Application Notes". The [Journal of Open Source Software (JOSS)](https://joss.theoj.org/)
offers programmers a way to generate a publication 
based on well-documented software;
in this process, the code itself is peer reviewed, and writing an actual paper is optional. For an example, see
@rosenbergTidyLPAPackageEasily2018.
Software citations can also sometimes be found 
in the source code repository (e.g., on Git remote repositories), using the
['Citation File Format'](https://citation-file-format.github.io/). The
[*Software Heritage Project*](https://www.softwareheritage.org/2020/05/26/citing-software-with-style/)
recently published a BibLaTeX software citation style that will further facilitate
accurate citation of software in dynamically generated documents.

@@

## Implementation in worcs

The \pkg{worcs} package includes a [vignette on citation](https://cjvanlissa.github.io/worcs/articles/citing.html) to explain the
fundamentals for novice users, and offer recommendations specific to \pkg{worcs}. We
recommend using the free, open-source reference manager
[Zotero](https://www.zotero.org/download/); it is feature-rich, 
user-friendly, and highly interoperable with other reference managers. A 
tutorial for using Zotero with RMarkdown exists  [here](https://christopherjunk.netlify.com/blog/2019/02/25/zotero-RMarkdown/).

The \pkg{worcs} package also offers original solutions for comprehensive citation.
Firstly,
during the writing process, authors can mark the distinction between essential
and non-essential references. It is easier to do this from the start, instead of
going back to cut non-essential references just prior to publication. Standard
Markdown uses the at-symbol (\code{@}) to cite a reference. \pkg{worcs} additionally 
reserves the "double at"-symbol (\code{@}) to cite a non-essential reference.
Users can render the manuscript either with, or without, comprehensive citations
by adapting the \code{knit} command in the front matter, setting it to
\code{knit: worcs::cite_all} to render all citations, and to 
\code{knit: worcs::cite_essential} to remove all *non-essential* citations.

With regard to the citation of \proglang{R} packages, it is worth noting that the \pkg{papaja} 
package [@austPapajaPrepareReproducible2020], which provides the \code{APA6}
template used in \pkg{worcs}, also
includes functions to automatically cite all packages used in the \proglang{R} session, and
add their references to the bibliography file.

@@

WORCS offers a workflow for open reproducible code in science. The step-by-step procedure outlined in this tutorial helps researchers make an entire research project Open and Reproducible. The accompanying \proglang{R} package provides user-friendly support functions for several steps in the workflow, and an RStudio project template to get the project set up correctly.
